Estimating initial state and quark-gluon plasma medium properties \\ using a hybrid model with nucleon substructure \\ calibrated to p-Pb and Pb-Pb collisions at \texorpdfstring{$\mathbf{\sqrts=5.02}$}{}~TeV

### 2018-08-06 paper/substructure.tex
Update numbers for latest posterior

Within the eikonal approximation, the initial entropy deposited at midrapidity at proper time $\tau=0^+\ \fmc$ can be represented by some mapping ${f(\T_A, \T_B) \mapsto dS/dy}$.
A natural first guess for the mapping is the arithmetic mean
\begin{equation}
  \frac{dS}{dy}\bigg\vert_{y=0} \propto \frac{\T_A + \T_B}{2}
\end{equation}
of the participant nuclear matter $\T_A$ and $T_B$, which is simply a wounded nucleon model up to meaningless factor of two in the normalization.
The wounded nucleon model was in fact one of the first such mappings used as a proxy for initial particle production and entropy deposition in relativistic heavy-ion collisions \cite{Bialas:1976ed}.
It was quickly realized, however, that the wounded nucleon model predicts the wrong scaling for charged particle production as a function of collision centrality and hence the wrong scaling for initial entropy deposition as a function of participant thicknesses $\T_A$ and $\T_B$ \cite{Kharzeev:2000ph}.

@@

\begin{table}[t]
  \caption{
    \label{tab:post_param}
    Posterior parameter estimates corresponding to Fig.~\ref{fig:posterior}.
    The reported values are for the distribution median and 90\% highest posterior density credible interval.
  }
  \begin{ruledtabular}
    \newlength{\cellwidth}
    \settowidth{\cellwidth}{$-0.00$}
    \newcommand{\est}[3]{\parbox{\cellwidth}{\hfill$#1$}$_{-#2}^{+#3}$}
    \begin{tabular}{llll}
      \toprule
      \multicolumn{2}{c}{Initial condition / Pre-eq}     & \multicolumn{2}{c}{QGP medium}              \\
      \cmidrule(r){1-2}                                    \cmidrule(l){3-4}
      \addlinespace[.4ex]
      Norm       & \est{20.3}{2.2}{2.2}          & $\eta/s$ min     & \est{0.10}{0.07}{0.07}            \\[1.1ex]
      $p$        & \est{0.018}{0.159}{0.146}     & $\eta/s$ slope   & \est{1.16}{1.16}{1.55} GeV$^{-1}$ \\[1.1ex]
      $\sigmaf$  & \est{0.93}{0.27}{0.27}        & $\eta/s$ crv     & \est{-0.21}{0.79}{0.85}           \\[1.1ex]
      $w$        & \est{0.97}{0.17}{0.19} fm     & $\zeta/s$ max    & \est{0.027}{0.027}{0.034}         \\[1.1ex]
      $n_c$      & \est{6.2}{3.1}{2.8}           & $\zeta/s$ width  & \est{0.032}{0.032}{0.047} GeV     \\[1.1ex]
      $\X$       & \est{0.34}{0.17}{0.22}        & $\zeta/s$ $T_0$  & \est{0.176}{0.020}{0.024} GeV     \\[1.1ex]
      $\dmin$    & \est{1.12}{0.50}{0.56} fm     & $\Tsw$           & \est{0.150}{0.012}{0.015} GeV     \\[1.1ex]
      $\taufs$   & \est{0.37}{0.27}{0.34} \fmc   & \\
      \addlinespace[.4ex]
      \bottomrule
    \end{tabular}
  \end{ruledtabular}
\end{table}

@@

The \trento\ normalization factor ${\text{Norm} = 20.3^{+2.2}_{-2.2}}$ and generalized mean parameter $p=0.018^{+0.146}_{-0.159}$ are well constrained by the present analysis.
Moreover, their posterior values nicely describe the $p$-Pb and Pb-Pb calibration observables in Figs.~\ref{fig:obs_pbpb} and \ref{fig:obs_ppb}.
While it would not be surprising, for example, to fit one or two of these observables using such a narrow range of values, the quality of the combined fit (more on this later) and the number of observables described is highly non-trivial.
For example, consider the ratio of the $p$-Pb yield over the Pb-Pb yield, which imposes a strong constraint on physically reasonable initial condition models.
As the entropy deposition parameter $p$ trends toward positive (negative) infinity, particle production scales like the local maximum (minimum) of the nuclear overlap density.
The parameter $p$ thus strongly affects the $p$-Pb and Pb-Pb yield ratio.
It just so happens that the small range of $p$-values needed to describe this yield ratio are the same values needed to describe all of the calibration observables in the present study, \emph{and} numerous Pb-Pb observables at $\sqrts=2.76$~TeV \cite{Bernhard:2018hnz}.
This work thus reaffirms an empirical scaling law reported by several previous studies \cite{Moreland:2014oya, Bernhard:2016tnd, Ke:2016jrd, Bernhard:2018hnz} for the initial transverse entropy density (or massless parton density) of an ultrarelativistic nuclear collision, given by
\begin{equation}
  \label{eq:geometric_mean}
  \frac{dS}{dy}\bigg\vert_{y=0} \propto \sqrt{\T_A \T_B},
\end{equation}
where $\T_A$ and $\T_B$ are the participant thickness functions \eqref{eq:fluctuated_thick} of each nucleus.
We emphasize that this specific analytic form should not be interpreted too literally.
For instance, a generalized mean described by $p=0.05$ is well within our 90\% credible interval, but it is not equal to the geometric mean in Eqn.~\ref{eq:geometric_mean}.
This approximate form nevertheless has been shown to mimic the scaling behavior of several saturation-based theory calculations in high-energy QCD \cite{Bernhard:2016tnd}, and thus appears consistent with theoretical expectations.

@@

Continuing down the diagonal in Fig.~\ref{fig:posterior}, we see that the nucleon width parameter $w=0.97_{-0.17}^{+0.19}$~fm.
We caution that this parameter $w$ is \emph{not} the RMS radius of our nucleons due to idiosyncrasies of our constituent sampling procedure.
We can, however, easily calculate the RMS radius for a specific nucleon width $w$, constituent width $v$ and constituent number $n_c$.
For example, the single highest posterior probability region of the parameter space prefers a nucleon width $w=0.92$~fm, constituent width $v=0.43$~fm and constituent number $n_c=6$.
The corresponding RMS nucleon radius for these parameters is $R_n = 0.86$~fm, conspicuously close the proton's RMS electric charge radius $R_p = 0.879(8)$~fm \cite{Bernauer:2010wm}.
This is perhaps the single largest difference between our work and the conclusions of recent saturation-based calculations which constrained the event-by-event fluctuations of the proton using a color-dipole picture of vector meson production \cite{Mantysaari:2016ykx, Mantysaari:2016jaz}.
Those studies find that the measured coherent and incoherent $J/\Psi$ spectra at HERA prefer a compact gluon distribution inside each nucleon, with an RMS radius $R_g \approx 0.4$~fm, which is roughly \emph{half} the analogous RMS nucleon width preferred by our analysis.

@@

The posterior on the nucleon substructure parameter $\X=0.34_{-0.17}^{+0.22}$, on the other hand, is particularly sharply peaked.
Recall that this parameter, defined in Eqn.~\eqref{eq:struct_param}, interpolates between the minimum and maximum widths of each constituent and hence
two different limits for the granularity of the nucleon.
When $\X=0$, the nucleon is populated by $n_c$ small compact hot-spots, and when $\X=1$ the hot-spots are large and fully overlapping, restoring spherical symmetry.

The nucleon structure parameter $\X$ is somewhat awkward to conceptualize, but we can easily transform its value back into a constituent width $v$.
Figure~\ref{fig:posterior_proton_shape} shows the resulting joint posterior distribution for the nucleon width parameter $w$ and constituent width $v$.
The posterior distribution, shown in blue, is remarkably well constrained.
We report a posterior estimate for the constituent width $v=0.46_{-0.15}^{+0.21}$~fm which is roughly the same magnitude as the \emph{nucleon} hot-spots used in recent saturation-based substructure studies that employed IP-Glasma initial conditions \cite{Schenke:2018fci}.
Evidently, it may be necessary to place an informative prior on our nucleon substructure parameters in order to resolve the apparent tension between our parameter values and those needed to describe DIS measurements at HERA.
Alternatively, it is also possible that the fluctuations probed by coherent and incoherent $J/\Psi$ production are different than those probed by minimum-bias particle production.

@@

The posterior free-streaming time $\taufs=0.37_{-0.27}^{+0.34}\ \fmc$ obtained in this work is significantly smaller than our previous estimate $\taufs=1.16_{-0.25}^{+0.29}\ \fmc$ in Ref.~\cite{Bernhard:2018hnz}.
We point out that the present study is missing several important observables which could affect the estimated free-streaming time, e.g.\ the Pb-Pb mean $p_T$ and mean $p_T$ fluctuations at $\sqrts=5.02$~TeV.
Nevertheless, it appears that the inclusion of nucleon substructure significantly reduces the maximum allowed free-streaming time, although more work is needed to establish if this is indeed the case.

@@

Second, we obtained a posterior distribution for the model input parameters in Fig.~\ref{fig:posterior} and reported quantitative estimates for these parameters in Tbl.~\ref{tab:post_param}.
Here we summarize our key findings about the model:
\begin{enumerate}[itemsep=0pt, leftmargin=2\parindent]
  \item
    Using \trento\ initial conditions, we find that initial entropy deposition (or particle production) scales like the geometric mean of local participant nuclear densities, in agreement with a number of previous estimates \cite{Moreland:2014oya, Bernhard:2016tnd, Bernhard:2018hnz}.
  \item
    Nucleon substructure is necessary to simultaneously describe $p$-Pb and Pb-Pb bulk observables. However, there is no strong preference for a specific number of constituents inside the nucleon.
  \item
    Bulk observables in $p$-Pb and Pb-Pb collisions prefer larger nucleons, similar in size to the proton's RMS electric charge radius.
    This agrees with a similar Bayesian analysis which was calibrated to Pb-Pb bulk observables at $\sqrts=2.76$ and 5.02~TeV using a hybrid model without nucleon substructure \cite{Bernhard:2018hnz}.
    We note that our result is in significant tension with an estimate for the effective nucleon width based on the gluon distribution extracted from HERA data \cite{Rezaeian:2012ji}.
  \item
    We obtain an estimate $v=0.46_{-0.15}^{+0.21}$~fm for the Gaussian width of the constituent hot spots inside each nucleon.
    This is significantly larger than the length scales typically associated with nucleon substructure, and it may help explain the seemingly unreasonable success of hydrodynamics in small collision systems.
  \item
    Our posterior estimate for the pre-equilibrium free-streaming time is $\taufs=0.37_{-0.27}^{+0.34}\ \fmc$.
    This is significantly shorter than the estimate obtained from a similar Bayesian study in Ref.~\cite{Bernhard:2018hnz}, which reported $\taufs=1.16_{-0.25}^{+0.29}\ \fmc$.
    It is not clear whether the difference is a result of nucleon substructure or the different observables used to calibrate each analysis.
  \item
    We compare in Fig.~\ref{fig:region_shear_bulk} our estimate for the temperature dependence of the QGP specific shear and bulk viscosities to those of Ref.~\cite{Bernhard:2018hnz}, which performed a Bayesian calibration to Pb-Pb bulk observables at $\sqrts=2.76$ and 5.02~TeV using a physics model without nucleon substructure.
    The two studies are in good agreement, although Ref.~\cite{Bernhard:2018hnz} obtains a more precise estimate for $(\eta/s)(T)$, likely due to the additional beam energies and observables included, and an enhanced sensitivity of larger collision systems to the QGP viscosity.
  \item
    We make predictions in Figs.~\ref{fig:obs_map} and \ref{fig:flow_corr} for several quantities which were not included in the model calibration, including the identified yields, transverse energy, symmetric cumulants, and mean $p_T$ fluctuations at $\sqrts=5.02$~TeV.
    Interestingly, our MAP estimate for the Pb-Pb symmetric cumulants at $\sqrts=5.02$~TeV are significantly different than those estimated in Ref.~\cite{Bernhard:2018hnz}.
    This could be a direct (or indirect) result of including nucleon substructure in the model calibration.
\end{enumerate}

>>>>

Within the eikonal approximation, the initial entropy deposited at midrapidity at proper time $\tau=0^+\ \fmc$ can be represented by some mapping ${f: \T_A, \T_B \mapsto dS/dy}$.
A natural first guess for the mapping is the arithmetic mean
\begin{equation}
  \frac{dS}{dy}\bigg\vert_{y=0} \propto \frac{\T_A + \T_B}{2}
\end{equation}
of the participant nuclear matter $\T_A$ and $T_B$, which is simply a wounded nucleon model up to meaningless factor of two in the normalization.
The wounded nucleon model was in fact one of the first such mappings used as a proxy for initial particle production and entropy deposition in relativistic heavy-ion collisions \cite{Bialas:1976ed}.
It was quickly realized, however, that the wounded nucleon model predicts the wrong scaling for charged particle production as a function of collision centrality and hence the wrong scaling for initial entropy deposition as a function of participant thicknesses $\T_A$ and $\T_B$ \cite{Kharzeev:2000ph}.

@@

\begin{table}[t]
  \caption{
    \label{tab:post_param}
    Posterior parameter estimates corresponding to Fig.~\ref{fig:posterior}.
    The reported values are for the distribution median and 90\% highest posterior density credible interval.
  }
  \begin{ruledtabular}
    \newlength{\cellwidth}
    \settowidth{\cellwidth}{$-0.00$}
    \newcommand{\est}[3]{\parbox{\cellwidth}{\hfill$#1$}$_{-#2}^{+#3}$}
    \begin{tabular}{llll}
      \toprule
      \multicolumn{2}{c}{Initial condition / Pre-eq}     & \multicolumn{2}{c}{QGP medium}              \\
      \cmidrule(r){1-2}                                    \cmidrule(l){3-4}
      \addlinespace[.4ex]
      Norm       & \est{20.2}{2.2}{2.2}          & $\eta/s$ min     & \est{0.10}{0.07}{0.07}            \\[1.1ex]
      $p$        & \est{0.013}{0.155}{0.144}     & $\eta/s$ slope   & \est{1.11}{1.11}{1.42} GeV$^{-1}$ \\[1.1ex]
      $\sigmaf$  & \est{0.93}{0.27}{0.26}        & $\eta/s$ crv     & \est{-0.14}{0.86}{0.81}           \\[1.1ex]
      $w$        & \est{0.98}{0.17}{0.18} fm     & $\zeta/s$ max    & \est{0.026}{0.026}{0.033}         \\[1.1ex]
      $n_c$      & \est{6.3}{3.1}{2.7}           & $\zeta/s$ width  & \est{0.033}{0.033}{0.048} GeV     \\[1.1ex]
      $\X$       & \est{0.35}{0.17}{0.22}        & $\zeta/s$ $T_0$  & \est{0.176}{0.020}{0.024} GeV     \\[1.1ex]
      $\dmin$    & \est{1.13}{0.51}{0.55} fm     & $\Tsw$           & \est{0.149}{0.013}{0.014} GeV     \\[1.1ex]
      $\taufs$   & \est{0.37}{0.27}{0.33} \fmc   & \\
      \addlinespace[.4ex]
      \bottomrule
    \end{tabular}
  \end{ruledtabular}
\end{table}

@@

The \trento\ normalization factor ${\text{Norm} = 20.2^{+2.2}_{-2.2}}$ and generalized mean parameter $p=0.013^{+0.144}_{-0.155}$ are well constrained by the present analysis.
Moreover, their posterior values nicely describe the $p$-Pb and Pb-Pb calibration observables in Figs.~\ref{fig:obs_pbpb} and \ref{fig:obs_ppb}.
While it would not be surprising, for example, to fit one or two of these observables using such a narrow range of values, the quality of the combined fit (more on this later) and the number of observables described is highly non-trivial.
For example, consider the ratio of the $p$-Pb yield over the Pb-Pb yield, which imposes a strong constraint on physically reasonable initial condition models.
As the entropy deposition parameter $p$ trends toward positive (negative) infinity, particle production scales like the local maximum (minimum) of the nuclear overlap density.
The parameter $p$ thus strongly affects the $p$-Pb and Pb-Pb yield ratio.
It just so happens that the small range of $p$-values needed to describe this yield ratio are the same values needed to describe all of the calibration observables in the present study, \emph{and} numerous Pb-Pb observables at $\sqrts=2.76$~TeV \cite{Bernhard:2018hnz}.
This work thus reaffirms an empirical scaling law reported by several previous studies \cite{Moreland:2014oya, Bernhard:2016tnd, Ke:2016jrd, Bernhard:2018hnz} for the initial transverse entropy density (or massless parton density) of an ultrarelativistic nuclear collision, given by
\begin{equation}
  \label{eq:geometric_mean}
  \frac{dS}{dy}\bigg\vert_{y=0} \propto \sqrt{\T_A \T_B},
\end{equation}
where $\T_A$ and $\T_B$ are the participant thickness functions \eqref{eq:fluctuated_thick} of each nucleus.
We emphasize that this specific analytic form should not be interpreted too literally.
For instance, a generalized mean described by $p=0.05$ is well within our 90\% credible interval, but it is not equal to the geometric mean in Eqn.~\ref{eq:geometric_mean}.
This approximate form nevertheless has been shown to mimic the scaling behavior of several saturation-based theory calculations in high-energy QCD \cite{Bernhard:2016tnd}, and thus appears consistent with theoretical expectations.

@@

Continuing down the diagonal in Fig.~\ref{fig:posterior}, we see that the nucleon width parameter $w=0.98_{-0.17}^{+0.18}$~fm.
We caution that this parameter $w$ is \emph{not} the RMS radius of our nucleons due to idiosyncrasies of our constituent sampling procedure.
We can, however, easily calculate the RMS radius for a specific nucleon width $w$, constituent width $v$ and constituent number $n_c$.
For example, the single highest posterior probability region of the parameter space prefers a nucleon width $w=0.92$~fm, constituent width $v=0.43$~fm and constituent number $n_c=6$.
The corresponding RMS nucleon radius for these parameters is $R_n = 0.86$~fm, conspicuously close the proton's RMS electric charge radius $R_p = 0.879(8)$~fm \cite{Bernauer:2010wm}.
This is perhaps the single largest difference between our work and the conclusions of recent saturation-based calculations which constrained the event-by-event fluctuations of the proton using a color-dipole picture of vector meson production \cite{Mantysaari:2016ykx, Mantysaari:2016jaz}.
Those studies find that the measured coherent and incoherent $J/\Psi$ spectra at HERA prefer a compact gluon distribution inside each nucleon, with an RMS radius $R_g \approx 0.4$~fm, which is roughly \emph{half} the analogous RMS nucleon width preferred by our analysis.

@@

The posterior on the nucleon substructure parameter $\X=0.35_{-0.17}^{+0.22}$, on the other hand, is particularly sharply peaked.
Recall that this parameter, defined in Eqn.~\eqref{eq:struct_param}, interpolates between the minimum and maximum widths of each constituent and hence
two different limits for the granularity of the nucleon.
When $\X=0$, the nucleon is populated by $n_c$ small compact hot-spots, and when $\X=1$ the hot-spots are large and fully overlapping, restoring spherical symmetry.

The nucleon structure parameter $\X$ is somewhat awkward to conceptualize, but we can easily transform its value back into a constituent width $v$.
Figure~\ref{fig:posterior_proton_shape} shows the resulting joint posterior distribution for the nucleon width parameter $w$ and constituent width $v$.
The posterior distribution, shown in blue, is remarkably well constrained.
We report a posterior estimate for the constituent width $v=0.47_{-0.15}^{+0.20}$~fm which is roughly the same magnitude as the \emph{nucleon} hot-spots used in recent saturation-based substructure studies that employed IP-Glasma initial conditions \cite{Schenke:2018fci}.
Evidently, it may be necessary to place an informative prior on our nucleon substructure parameters in order to resolve the apparent tension between our parameter values and those needed to describe DIS measurements at HERA.
Alternatively, it is also possible that the fluctuations probed by coherent and incoherent $J/\Psi$ production are different than those probed by minimum-bias particle production.

@@

The posterior free-streaming time $\taufs=0.37_{-0.27}^{+0.33}\ \fmc$ obtained in this work is significantly smaller than our previous estimate $\taufs=1.16_{-0.25}^{+0.29}\ \fmc$ in Ref.~\cite{Bernhard:2018hnz}.
We point out that the present study is missing several important observables which could affect the estimated free-streaming time, e.g.\ the Pb-Pb mean $p_T$ and mean $p_T$ fluctuations at $\sqrts=5.02$~TeV.
Nevertheless, it appears that the inclusion of nucleon substructure significantly reduces the maximum allowed free-streaming time, although more work is needed to establish if this is indeed the case.

@@

Second, we obtained a posterior distribution for the model input parameters in Fig.~\ref{fig:posterior} and reported quantitative estimates for these parameters in Tbl.~\ref{tab:post_param}.
Here we summarize our key findings about the model:
\begin{enumerate}[itemsep=0pt, leftmargin=2\parindent]
  \item
    Using \trento\ initial conditions, we find that initial entropy deposition (or particle production) scales like the geometric mean of local participant nuclear densities, in agreement with a number of previous estimates \cite{Moreland:2014oya, Bernhard:2016tnd, Bernhard:2018hnz}.
  \item
    Nucleon substructure is necessary to simultaneously describe $p$-Pb and Pb-Pb bulk observables. However, there is no strong preference for a specific number of constituents inside the nucleon.
  \item
    Bulk observables in $p$-Pb and Pb-Pb collisions prefer larger nucleons, similar in size to the proton's RMS electric charge radius.
    This agrees with a similar Bayesian analysis which was calibrated to Pb-Pb bulk observables at $\sqrts=2.76$ and 5.02~TeV using a hybrid model without nucleon substructure \cite{Bernhard:2018hnz}.
    We note that our result is in significant tension with an estimate for the effective nucleon width based on the gluon distribution extracted from HERA data \cite{Rezaeian:2012ji}.
  \item
    We obtain an estimate $v=0.47_{-0.15}^{+0.20}$~fm for the Gaussian width of the constituent hot spots inside each nucleon.
    This is significantly larger than the length scales typically associated with nucleon substructure, and it may help explain the seemingly unreasonable success of hydrodynamics in small collision systems.
  \item
    Our posterior estimate for the pre-equilibrium free-streaming time is $\taufs=0.37_{-0.27}^{+0.33}\ \fmc$.
    This is significantly shorter than the estimate obtained from a similar Bayesian study in Ref.~\cite{Bernhard:2018hnz}, which reported $\taufs=1.16_{-0.25}^{+0.29}\ \fmc$.
    It is not clear whether the difference is a result of nucleon substructure or the different observables used to calibrate each analysis.
  \item
    We compare in Fig.~\ref{fig:region_shear_bulk} our estimate for the temperature dependence of the QGP specific shear and bulk viscosities to those of Ref.~\cite{Bernhard:2018hnz}, which performed a Bayesian calibration to Pb-Pb bulk observables at $\sqrts=2.76$ and 5.02~TeV using a physics model without nucleon substructure.
    The two studies are in good agreement, although Ref.~\cite{Bernhard:2018hnz} obtains a more precise estimate for $(\eta/s)(T)$, likely due to the additional beam energies and observables included, and an enhanced sensitivity of larger collision systems to the QGP viscosity.
  \item
    We make predictions in Figs.~\ref{fig:obs_map} and \ref{fig:flow_corr} for several quantities which were not included in the model calibration, including the identified yields, transverse energy, symmetric cumulants, and mean $p_T$ fluctuations at $\sqrts=5.02$~TeV.
    Interestingly, our MAP estimate for the Pb-Pb symmetric cumulants at $\sqrts=5.02$~TeV are significantly different than those estimated in Ref.~\cite{Bernhard:2018hnz}.
    This could be a direct (or indirect) result of including nucleon substructure in the model calibration.
\end{enumerate}
